Relevance-based Quantization of Scattering Features for Unsupervised Mining of Environmental Audio

### 2018-02-17 paper/structureScene.tex
Edit introduction

The amount of audio data recorded from our sonic environment has grown considerably over the past decades.
In order to measure the effect of human activity and climate change on animal biodiversity, researchers have recently undertaken a massive deployment of acoustic sensors throughout the world \cite{warren2006urban, NessSST13, stowell13b}.
In addition, recent work has explored acoustic monitoring for characterization of human pleasantness in urban areas \cite{guyot2005urban, ricciardi2015sound}, as well as the prediction of annoyance due to traffic \cite{gloaguen}.
Since they bear a strong societal impact and raise many scientific challenges, we believe that these applications are of considerable interest for the signal processing community.

This work falls in the realm of ecoacoustics, an emerging field in computational bioacoustics, which is the interdisciplinary study of relationships between natural or anthropogenic sounds and their environments at the intersection of ecology, acoustics and computer science \cite{krause}. %Despite identified needs, the field of signal-based ecoacoustics is still in its infancy.
%Consequently, few well-designed datasets are readily available for evaluation purposes.
An important problem in this field is that manually analysing the recorded data to identify the quantities of interest is very costly \cite{wimmer2013sampling}. Some sort of pre-screening is therefore needed to reduce the need for human expert listening and annotation.

The most straightforward approach is to assume a closed set of events of interests like sounds of animals expected to live near the acoustic sensors, train models for them and perform automatic annotation \cite{}. A given time interval, for example a single day, is then represented by the number of events within the predefined typology that have been detected in that period of time. This approach allows the scientist to drastically reduce the amount of information that needs to be processed. However, this approach has two drawbacks. The first is the reliance to trained models whose behavior on unseen data (such as different sensors) is prone to errors and cannot be always be trusted. More importantly, it is based on \textit{a priori} knowledge and thus cannot be considered for exploratory analysis where the quantities of interest have yet to be defined.

In order to identify which parts needs human inspection, one needs tools that can mine data and annotate to whether recurring patterns  or sparsely distributed events to study anomalous or scarce behaviors. The latter allowing the scientist to then annotate only a representative audio of the periods of time with recurring patterns and the former to potentially discover unforseen phenomena.

With this aim, one needs to design an algorithm to perform an acoustic similarity retrieval, where the audio fragments judged ``most similar'' to a given query section are extracted from some larger dataset. This requires the audio fragments to be represented in some way that captures their distinctive qualities. This has been achieved by bag-of-frames approaches \cite{aucouturier2007bag}, which attempt to describe the auditory scene recording using summary statistics. Unfortunately, since this only captures the average structure of the scene, the approach often fails when presented with scenes which are dynamic and contain episodic, distinct sound events that discriminate the scene from others. Indeed, experiments in cognitive psychology \cite{dubois2006cognitive} and cognitive neurosciences \cite{nelken2004processing} suggest that this matches more closely human acoustic perception of acoustic scenes. We believe that the failure to model such distinct events is one of the reason why this approach is found to be insufficient when considering evaluation datasets of typical sizes \cite{lagrange:hal-01082501}.

%To design better representations for the similarity retrieval task, we draw inspiration from a closely related, and more mature, field of investigation: the classification of acoustic scenes. This task, consisting of predicting scene labels from audio recordings, is narrower in its range of applications, but has the advantage of being rooted in the cognitive psychology of categorization \cite{dubois2006cognitive,guastavino_ideal_2006}. In addition, it has received significant attention from the data mining community for some time, resulting in the availability of well-designed datasets and competitive algorithms for solving it.

Addressing this task first requires the ability to capture meaningful aspects within the signal at small time-scales.
In other words, how do we represent an extract of the auditory scene signal over a sub-second interval?
A standard tool for this are the mel-frequency cepstral coefficients (MFCCs), originally developed for speech processing \cite{davis-mermelstein}, but which have found wider use in recent years in music information retrieval \cite{logan} and environmental audio processing \cite{aucouturier2007bag}.
A richer set of features is found in the scattering transform, a generic signal representation which has enjoyed significant success in various audio classification tasks \cite{Anden2014,chudacek}.
This is because the scattering transform cascades wavelet convolutions and pointwise nonlinearities, thus imitating the architecture of a convolutional neural network \cite{lee,lostanlen-deep-spiral,soundnet,arandjelovic-zisserman}.

For the task at at hand, scattering features have the advantage of not requiring training and thus potentially lead to wider range of application than trained features. Indeed, the task at hand is the mining of yet unheard audio datasets and the properties of the audio that are of interest remains to be defined, thus leading to an unserpervised setting.

%The next question is the choice of a classifier. Here, a class of particularly successful algorithms train a classifier, such as a support vector machine (SVM) \cite{cristianini2000introduction}, to discriminate between short extracts of an auditory scene, and at the classification stage divide the signal into short extracts which are classified separately before pooling the classification results to reach a final decision. This last step is known as late integration and is crucial to the success of these methods, since it allows the classifier to concentrate on the short, distinct sound events that discriminate between scenes. The success of the late integration technique again leads credence to the idea that rare, distinct events in an auditory scene are sufficient to characterize it almost completely. Combining scattering transforms for low-level representation with late-integrated SVMs for classifying the scene at a higher level, we obtain a competitive auditory scene classification system, demonstrating the usefulness of both concepts for this task.

%we can emulate their selection behavior by quantizing the auditory scene and defining a distance over this quantization.
We also propose a new model for acoustic scenes, where the signal is represented at sub-second scales by scattering transforms, while larger scales are captured by an unsupervised method that quantizes the scattering coefficients into a given number of clusters. We then use these clusters to define a set of distances that can subsequently be used for similarity retrieval. \vl{The verb use is repeated in the previous sentence.} Testing this approach on a scene retrieval task, we obtain significant improvements over traditional bag-of-frames and summary statistics models applied both to MFCCs and scattering coefficients.

Motivations of the proposed approach and a brief review of the state of the art in acoustic scene modeling are given in Section \ref{sec:soa}. We describe the scattering transform in Section \ref{sec:scattering}, discuss feature post-processing in Section \ref{sec:design} and propose a cluster-based scene description in Section \ref{sec:object}. Section \ref{sec:experiments} describes several experiments for the acoustic scene similarity retrieval task. Results are subsequently reported in Section \ref{sec:results}.

>>>>

The amount of audio data recorded from our sonic environment has grown considerably over the past decades.
In order to measure the effect of human activity and climate change on animal biodiversity, researchers have recently undertaken a massive deployment of acoustic sensors throughout the world \cite{warren2006urban, NessSST13, stowell13b}.
In addition, recent work has explored acoustic monitoring for characterization of human pleasantness in urban areas \cite{guyot2005urban, ricciardi2015sound}, as well as the prediction of annoyance due to traffic \cite{gloaguen}.
Since they bear a strong societal impact and raise many scientific challenges, we believe that these applications are of considerable interest for to signal processing community.

These questions are explored in an emerging field of computational bioacoustics, known as ecoacoustics, which is the interdisciplinary study of relationships between natural or anthropogenic sounds and their environments at the intersection of ecology, acoustics and computer science \cite{krause}.
An important problem in this field is that manually analysing the recorded data to identify the quantities of interest is very costly \cite{wimmer2013sampling}. Some sort of pre-screening is therefore required to reduce the need for human expert listening and annotation.

The most straightforward approach is to specify a closed set of sound classes, such as sounds of animals expected to live near the acoustic sensors. We then train models for these classes which are used to automatically annotate our recordings \cite{}. \ja{What's the reference here?} A given time interval (e.g., a single day) is then represented by the number of events detected during that interval for each class. This allows the scientist to drastically reduce the amount of information requiring manual processing. However, this approach has two drawbacks. First, it relies on trained models whose behavior on unseen data (such as different sensors) is prone to errors and cannot be trusted. Second, and more importantly, it is based on \textit{a priori} knowledge and thus cannot be considered for exploratory analysis where quantities of interest have yet to be defined.

To identify which parts need human inspection, we need tools that can detect both recurring patterns and sparsely distributed events. Identifying recurring patterns allows the user to focus on certain time points for manual annotation, while detection of more rare structures enables discovery of unforeseen phenomena.

With this aim, we need to design an algorithm for acoustic similarity retrieval, where the audio fragments judged ``most similar'' to a given query recording must be extracted from some larger dataset. This requires us to represent an audio recording in a way that captures its distinctive qualities. This has previously been attempted using bag-of-frames approaches \cite{aucouturier2007bag}, which describes an auditory scene recording using summary statistics of short-time features. Unfortunately, this only captures the average structure of the scene, so the approach often fails when presented with highly dynamic scenes or those characterized by a few distinct sound events. Indeed, experiments in cognitive psychology \cite{dubois2006cognitive} and cognitive neuroscience \cite{nelken2004processing} suggest that this more closely matches human acoustic perception. We believe that the failure to model such distinct events is one of the reasons why this approach is insufficient for moderate-sized evaluation datasets \cite{lagrange:hal-01082501}.

Solving the acoustic similarity retrieval first requires the ability to capture meaningful signal structure at small time scales.
This is often achieved using mel-frequency cepstral coefficients (MFCCs).
Originally developed for speech processing \cite{davis-mermelstein}, MFCCs have recently found wider use in music information retrieval \cite{logan} and environmental audio processing \cite{aucouturier2007bag}.
A richer representation, the scattering transform, has enjoyed significant success in various audio \cite{Anden2014} and biomedical \cite{chudacek} signal classification tasks.
Its structure is that of a convolutional neural network \cite{lee,lostanlen-deep-spiral,soundnet,arandjelovic-zisserman}, but with fixed filters.
Specifically, it alternates convolutions with wavelet filters and pointwise nonlinearities to ensure time-shift invariance and time-warping stability \cite{Mallat2012}.

For our task, one advantage of the scattering transform is not requiring a training step, allowing for a wider range of applications compared to learned features. Indeed, for data mining of previously unheard datasets, the properties of relevant audio structures remains to be defined, leading to an unserpervised setting.

In this work, we propose a new model for acoustic scenes, where the signal is represented at sub-second scales by scattering transforms, while larger scales are captured by an cluster model. This unsupervised model quantizes the scattering coefficients into a given number of clusters. These clusters to define a set of distances for acoustic similarity retrieval. Evaluating this approach on a scene retrieval task, we obtain significant improvements over traditional bag-of-frames and summary statistics models applied both to MFCCs and scattering coefficients.

Motivations of the proposed approach and a brief review of the state of the art in acoustic scene modeling are given in Section \ref{sec:soa}. We describe the scattering transform in Section \ref{sec:scattering}, discuss feature post-processing in Section \ref{sec:design} and propose a cluster-based scene description in Section \ref{sec:object}. Section \ref{sec:experiments} describes several experiments for the acoustic scene similarity retrieval task. Results are reported in Section \ref{sec:results}.
